---
title: "Assignment9"
author: "Stephen Ling"
output: html_document
---

```{r}
knitr::opts_chunk$set(echo = TRUE)
library(devtools)
install_github("genomicsclass/dagdata")
library(dagdata)
data(admissions)
install_github("genomicsclass/GSE5859")
library(GSE5859)
data(GSE5859)
```


#Batch Effect Exercise 1. 

1.
```{r}
index = which(admissions$Gender==1)
accepted= sum(admissions$Number[index] * admissions$Percent[index]/100)
applied = sum(admissions$Number[index])
accepted/applied
    
index = which(admissions$Gender==0)
accepted= sum(admissions$Number[index] * admissions$Percent[index]/100)
applied = sum(admissions$Number[index])
accepted/applied
```

2.
```{r}
atab<-as.data.frame(matrix(0,2,2))
rownames(atab)<-c("Men","Women")
colnames(atab)<-c("Accepted","Rejected")
index = admissions$Gender==1
men=admissions[index,]
women=admissions[!index,]
atab[1,1]= sum(men$Number * men$Percent/100)
atab[1,2]= sum(men$Number*(1-men$Percent/100))
windex = which(admissions$Gender==0)
atab[2,1]= sum(women$Number * women$Percent/100)
atab[2,2]= sum(women$Number * (1-women$Percent/100))
atab
chisq.test(atab)$p.value
```

How can this be? This is referred to as Simpson’s Paradox. In the following questions we will try to decipher why this is happening.

```{r}
index = admissions$Gender==1
men = admissions[index,]
women = admissions[!index,]
print( data.frame( major=admissions[1:6,1],men=men[,3], women=women[,3]) )
```

3.
```{r}
major = admissions[1:6,1]
men = admissions[1:6,]
women =admissions[7:12,]
H = (men$Number*men$Percent/100 + women$Number*women$Percent/100) / (men$Number+women$Number)
H
major[which.min(H)]
min(H)
```

Major F is the hardest.

4.
```{r}
min(H)
```

5. 
```{r}
cor(H,men$Number)
cor(men$Number,H)
```

The proportion about 0.765.

6. 
```{r}
cor(H,women$Number)
```

7.
There is a counfounding between gender and preference for "harder" majors: females are more likely to apply to harder majors.

###Batch Effect Exercises 2

```{r}
library(Biobase)
library(GSE5859)
data(GSE5859)
```

1.
```{r}
sampleInfo <- pData(e)
year = format(sampleInfo$date,"%y")
```

```{r}
table(year,sampleInfo$ethnicity)
```

2. 
```{r}
tab=table(year,sampleInfo$ethnicity)
print(tab)
x=rowSums(tab!=0)
sum(x>=2)
```
Two years. 

3.
```{r}
month.year = format(sampleInfo$date,"%m%y")
```

For what proportion of these month.year values do we have more than one ethnicity
represented?

```{r}
mytab=table(month.year,sampleInfo$ethnicity)
print(mytab)
myx=rowSums(mytab!=0)
mean(myx>=2)
```
The proportion is 0.047619

4.
```{r}
library(qvalue)
library(genefilter)
geneExpression = exprs(e)
year = factor( format(sampleInfo$date,"%y") )
index = which(year%in% c("02","03") & sampleInfo$ethnicity=="CEU")
year = droplevels(year[index])
pval = rowttests(geneExpression[ ,index], year)$p.value
qval = qvalue(pval)
sum(qval$qvalue < 0.05)
```

5.
```{r}
qval$pi0
```

6. 
```{r}
library(qvalue)
library(genefilter)
year = factor( format(sampleInfo$date,"%y") )
index = which(year%in% c("03","04") & sampleInfo$ethnicity=="CEU")
year = droplevels(year[index])
pval = rowttests(geneExpression[ ,index], year)$p.value
qval = qvalue(pval)
sum(qval$qvalue < 0.05)
```

7.
```{r}
library(qvalue)
library(genefilter)
ethnicity = factor( sampleInfo$ethnicity)
index = which(ethnicity%in% c("CEU","ASN"))
ethnicity = droplevels(ethnicity[index])
pval = rowttests(geneExpression[ ,index], ethnicity)$p.value
qval = qvalue(pval)
sum(qval$qvalue < 0.05)
```

8. 
```{r}
library(qvalue)
library(genefilter)
ethnicity = factor( sampleInfo$ethnicity)
year = factor( format(sampleInfo$date,"%y") )
index = which(ethnicity%in% c("CEU","ASN") & year=="05")
ethnicity = droplevels(ethnicity[index])
pval = rowttests(geneExpression[ ,index], ethnicity)$p.value
qval = qvalue(pval)
sum(qval$qvalue < 0.05)
```

```{r}
table(sampleInfo$ethnicity[index])
```

9.
```{r}
library(qvalue)
library(genefilter)
ethnicity = factor( sampleInfo$ethnicity)
year = factor( format(sampleInfo$date,"%y") )
indexASN = which(ethnicity%in% c("ASN") & year=="05")
indexCEU = which(ethnicity%in% c("CEU") & year=="02")
set.seed(3)
indexCEU<-sample(indexCEU,3)
index<-c(indexASN,indexCEU)
ethnicity = droplevels(ethnicity[index])
pval = rowttests(geneExpression[ ,index], ethnicity)$p.value
qval = qvalue(pval)
sum(qval$qvalue < 0.05)
```

#Batch Effect Exercise 3

```{r}
sex = sampleInfo$group
month = factor( format(sampleInfo$date,"%m"))
table( sampleInfo$group, month)
```

###1. 
```{r}
library(qvalue)
library(genefilter)
sex = factor( sex)
pval = rowttests(geneExpression, sex)$p.value
qval = qvalue(pval)
sum(qval$qvalue < 0.1)
```

###2. 
```{r}
library(qvalue)
library(genefilter)
sex = factor( sex)
pval = rowttests(geneExpression, sex)$p.value
qval = qvalue(pval)
qvals = qval$qvalues
index = geneAnnotation$CHR[qvals<0.1]%in%c("chrX","chrY")
mean(index)
```

###3.
```{r}
```

###4. 
```{r}
library(qvalue)
library(genefilter)
sex = factor( sex)
pval = rowttests(geneExpression, sex)$p.value
qval = qvalue(pval)
qvals = qval$qvalues
index = which(qvals<0.1 & !geneAnnotation$CHR%in%c("chrX","chrY"))
month = factor( format(sampleInfo$date,"%m"))
pval = rowttests(geneExpression[index,], month)$p.value
mean(pval<0.05)
```

###5.
The answer is D.

###6. 
```{r}
X = model.matrix(~sex+month)
i = 234
y = geneExpression[i,]
fit = lm(y~X)
summary(fit)$coef

pvals <- t( sapply(1:nrow(geneExpression),function(j){
  y <- geneExpression[j,]
  fit <- lm(y~X-1)
  summary(fit)$coef[2,4] # the 2 picks out the second (sex) coef, the 4 picks out the p value
} ) )
library(qvalue)
library(genefilter)
qval = qvalue(pvals)
qvals = qval$qvalues
sum(qvals<0.1)
```

###7. 
```{r}
index = geneAnnotation$CHR[qvals<0.1]%in%c("chrX","chrY")
mean(index)
```

###8. 
```{r}
```

###9. 
```{r}
X = model.matrix(~sex+month)
pvals <- t( sapply(1:nrow(geneExpression),function(j){
  y <- geneExpression[j,]
  fit <- lm(y~X-1)
  summary(fit)$coef[3,4] # third coef, for month
} ) )
library(qvalue)
library(genefilter)
qval = qvalue(pvals)
qvals = qval$qvalues
sum(qvals<0.1)
```

#Batch Effect Exercise #4

###1. 
The answer is B. 

###2. 
The answer is C.

###3. 
```{r}
library(Biobase)
library(GSE5859Subset)
data(GSE5859Subset)

y = geneExpression - rowMeans(geneExpression)

library(RColorBrewer)
library(rafalib)
mypar(1,2)
n=ncol(y)
cors=cor(y)
cols=colorRampPalette(rev(brewer.pal(11,"RdBu")))(100)
image(1:n,1:n,cors,xaxt="n",yaxt="n",col=cols,xlab="",ylab="",zlim=c(-1,1))
o <- order(sampleInfo$date)
yo<-y[,o]
cors=cor(yo)
cols=colorRampPalette(rev(brewer.pal(11,"RdBu")))(100)
image(1:n,1:n,cors,xaxt="n",yaxt="n",col=cols,xlab="",ylab="",zlim=c(-1,1))

library(rafalib)
sex = sampleInfo$group
mypar(1,2)
cors = cor(y)
image(cors)
o = order(sampleInfo$date)
image(cors[o,o])
```

###4.
```{r}
s <- svd(y)
What <- s$v[,1:2] # these are the first two PCs.
rownames(What)<-colnames(y)
round(What,2)
fit = s$u[,1:2]%*% (s$d[1:2]*t(What))
var(as.vector(fit))/var(as.vector(y))
```

###5.
```{r}
month=format(sampleInfo$date, "%m")
pcs = svd(y)$v[,1:2]
o = order(sampleInfo$date)
cols = as.numeric(month)[o]
mypar(2,1)
for(i in 1:2){
  plot(pcs[o,i],col=cols,xaxt="n",xlab="")
  label = gsub("2005-","",sampleInfo$date[o])
  axis(1,1:ncol(y),label,las=2)
```

The answer is A.

###6.
```{r}
s = svd(y)
varexplained = s$d^2/ sum(s$d^2)
plot(varexplained)
sum(varexplained>0.10)
```

###7/8.
```{r}
s = svd(y)
month = factor( format(sampleInfo$date,"%m"))
cors = cor( as.numeric(month),s$v)
plot(t(cors))
which.max(abs(cors))
max(abs(cors))
```

###9/10.
```{r}
s = svd(y)
sex = sampleInfo$group
cors = cor( as.numeric(sex),s$v)
plot(t(cors))
which.max(abs(cors))
max(abs(cors))
```

###11.
```{r}
X <- model.matrix(~sex+s$v[,1:2])

library(limma)
X <- model.matrix(~sex+s$v[,1:2])
fit <- lmFit(y,X)
k <- 2 ##second coef, for month
ses <- fit$stdev.unscaled[,k]*fit$sigma
ttest <- fit$coef[,k]/ses
pvals <- 2*pt(-abs(ttest),fit$df)
library(qvalue)
library(genefilter)
qval = qvalue(pvals)
qvals = qval$qvalues
sum(qvals<0.1)
```

###12.
```{r}
index = geneAnnotation$CHR[qvals<0.1]%in%c("chrX","chrY")
mean(index)
```

#Batch Effect Exercise #5

###1.
```{r}
library("sva")
s <- svd(geneExpression-rowMeans(geneExpression))
cor(sampleInfo$group,s$v[,1])

sex = sampleInfo$group
mod = model.matrix(~sex)
svafit = sva(geneExpression,mod)
head(svafit$sv)

for(i in 1:ncol(svafit$sv)){
  print( cor(s$v[,i],svafit$sv[,i]) )
}

svaX<-model.matrix(~sex+svafit$sv)
library(limma)
fit <- lmFit(geneExpression,svaX)
k <- 2 #second coef, for sex
ses <- fit$stdev.unscaled[,k]*fit$sigma
ttest <- fit$coef[,k]/ses
pvals <- 2*pt(-abs(ttest),fit$df)
library(qvalue)
library(genefilter)
qval = qvalue(pvals)
qvals = qval$qvalues
sum(qvals<0.1)
```

###2.
```{r}
index = geneAnnotation$CHR[qvals<0.1]%in%c("chrX","chrY")
mean(index)

res = sapply(1:nrow(geneExpression),function(i){
	y = geneExpression[i,]
	fit = lm(y~svaX-1)
	summary(fit)$coef[2,c(1,4)]
})
qvals = qvalue(res[2,])$qvalue
pcutoff = max( res[2,qvals < .1] )
library(rafalib)
mypar(1,1)
plot(res[1,],-log10(res[2,]),xlab="M",ylab="log10 p-value")
ind = which(geneAnnotation$CHR=="chrY")
points(res[1,ind],-log10(res[2,ind]),col=1,pch=16)
ind = which(geneAnnotation$CHR=="chrX")
points(res[1,ind],-log10(res[2,ind]),col=2,pch=16)
abline(h=-log10(pcutoff))
legend("bottomleft",c("chrX","chrY"),col=c(2,1),pch=16)